---
title: "timings_R_h2o"
author: "David Cortes"
date: "10/1/2021"
output:
  html_document:
    toc: true
    theme: united
---

# Timing Isolation Forest in H2O library

This is an extension of the timings from different isolation forest implementations in R from [here](https://github.com/david-cortes/isotree/blob/master/timings/timings_R.Rmd), this time evaluating the H2O library (version 3.34.0.1 at the time of writing). For more details, see the link above and the [GitHub repository](https://www.github.com/david-cortes/isotree) where this came from.

Note that this library always runs multi-threaded, thus there was no separate single-threaded comparison like for the other libraries.

```{r, message=FALSE, results=FALSE, warning=FALSE, comment=FALSE}
library(R.matlab) #https://github.com/HenrikBengtsson/R.matlab
library(readsparse) #https://github.com/david-cortes/readsparse
library(h2o) #https://docs.h2o.ai/h2o/latest-stable/h2o-docs/index.html
library(microbenchmark) #https://github.com/joshuaulrich/microbenchmark/
h2o.no_progress()
h2o.init()
```

### Small dataset: Satellite (6435 rows, 36 columns)

```{r}
X = readMat("satellite.mat")$X
X = as.h2o(X)
cat(dim(X))
```

```{r}
microbenchmark({
	iso = h2o.isolationForest(X, names(X),
							  ntrees=100L, max_depth=8L, sample_size=256L,
							  seed=1L)
}, times=7L)
```

```{r}
microbenchmark({
	iso = h2o.isolationForest(X, names(X),
							  ntrees=100L, max_depth=10L, sample_size=1024L,
							  seed=1L)
}, times=7L)
```

```{r}
microbenchmark({
	iso = h2o.isolationForest(X, names(X),
							  ntrees=100L, max_depth=13L, sample_size=6435L,
							  seed=1L)
}, times=7L)
```


```{r}
microbenchmark({
	iso = h2o.extendedIsolationForest(X, names(X), extension_level=1,
									  ntrees=100L, sample_size=256L,
									  seed=1L)
}, times=7L)
```

```{r}
microbenchmark({
	iso = h2o.extendedIsolationForest(X, names(X), extension_level=1,
									  ntrees=100L, sample_size=1024L,
									  seed=1L)
}, times=7L)
```

```{r}
microbenchmark({
	iso = h2o.extendedIsolationForest(X, names(X), extension_level=1,
									  ntrees=100L, sample_size=6435L,
									  seed=1L)
}, times=7L)
```


### Mid-sized dataset: CovType (581,012 rows, 54 columns)

```{r, message=FALSE, results=FALSE, warning=FALSE, comment=FALSE}
gc()
```
```{r}
X = readsparse::read.sparse("covtype.libsvm.binary")
X = as.matrix(X$X)
X = as.h2o(X)
cat(dim(X))
```

```{r}
microbenchmark({
	iso = h2o.isolationForest(X, names(X),
							  ntrees=100L, max_depth=8L, sample_size=256L,
							  seed=1L)
}, times=7L)
```
```{r}
microbenchmark({
	iso = h2o.isolationForest(X, names(X),
							  ntrees=100L, max_depth=10L, sample_size=1024L,
							  seed=1L)
}, times=7L)
```
```{r}
microbenchmark({
	iso = h2o.isolationForest(X, names(X),
							  ntrees=100L, max_depth=14L, sample_size=10000L,
							  seed=1L)
}, times=7L)
```


```{r}
microbenchmark({
	iso = h2o.extendedIsolationForest(X, names(X), extension_level=1,
									  ntrees=100L, sample_size=256L,
									  seed=1L)
}, times=7L)
```
```{r}
microbenchmark({
	iso = h2o.extendedIsolationForest(X, names(X), extension_level=1,
									  ntrees=100L, sample_size=1024L,
									  seed=1L)
}, times=7L)
```
```{r}
microbenchmark({
	iso = h2o.extendedIsolationForest(X, names(X), extension_level=1,
									  ntrees=100L, sample_size=10000L,
									  seed=1L)
}, times=7L)
```
