---
title: "timings_R"
author: "David Cortes"
date: "2/2/2021"
output:
  html_document:
    toc: true
    theme: united
---

# Timing Isolation Forest libraries

This document produces timings on fitting Isolation Forest (iForest) models to 2 datasets of different sizes using the libraries [IsoTree](https://github.com/david-cortes/isotree), [IsolationForest](https://sourceforge.net/projects/iforest/), [IsoFor](https://github.com/Zelazny7/isofor), and [Solitude](https://github.com/talegari/solitude).

All of the models produce 100 trees of the same fixed sample sizes. Some are run single-threaded and some multi-threaded where supported. The CPU used is an AMD Ryzen 2 2700 running at 3.2GHz and having 16 threads.

The second dataset was not benchmarked for all libraries because some of them would run out of memory and/or take a very unreasonable amount of time to finish.

******

Libraries used throughout the document:
```{r, message=FALSE, results=FALSE, warning=FALSE, comment=FALSE}
library(R.matlab) #https://github.com/HenrikBengtsson/R.matlab
library(readsparse) #https://github.com/david-cortes/readsparse
library(microbenchmark) #https://github.com/joshuaulrich/microbenchmark/
library(isotree) #https://github.com/david-cortes/isotree
library(IsolationForest) #https://sourceforge.net/projects/iforest
library(isofor) #https://github.com/Zelazny7/isofor
library(solitude) #https://github.com/talegari/solitude
```

### Small dataset: Satellite (6435 rows, 36 columns)

Data was taken from the ODDS repository - [link](http://odds.cs.stonybrook.edu/satellite-dataset/).


```{r}
X = readMat("satellite.mat")$X
X_df = as.data.frame(X)
cat(dim(X))
```

******

Timing isotree, single-variable model:
```{r}
microbenchmark({
	iso = isotree::isolation.forest(X, ntrees=100L, sample_size=256L,
									ndim=1, missing_action="fail",
									nthreads=1L)
}, times=7L)
```

```{r}
microbenchmark({
	iso = isotree::isolation.forest(X, ntrees=100L, sample_size=1024L,
									ndim=1, missing_action="fail",
									nthreads=1L)
}, times=7L)
```

```{r}
microbenchmark({
	iso = isotree::isolation.forest(X, ntrees=100L, sample_size=6435L,
									ndim=1, missing_action="fail",
									nthreads=1L)
}, times=7L)
```

```{r}
microbenchmark({
	iso = isotree::isolation.forest(X, ntrees=100L, sample_size=256L,
									ndim=1, missing_action="fail",
									nthreads=parallel::detectCores())
}, times=7L)
```

```{r}
microbenchmark({
	iso = isotree::isolation.forest(X, ntrees=100L, sample_size=1024L,
									ndim=1, missing_action="fail",
									nthreads=parallel::detectCores())
}, times=7L)
```

```{r}
microbenchmark({
	iso = isotree::isolation.forest(X, ntrees=100L, sample_size=6435L,
									ndim=1, missing_action="fail",
									nthreads=parallel::detectCores())
}, times=7L)
```

Timing isotree, extended model:
```{r}
microbenchmark({
	iso = isotree::isolation.forest(X, ntrees=100L, sample_size=256L,
									ndim=2, missing_action="fail",
									nthreads=1L)
}, times=7L)
```

```{r}
microbenchmark({
	iso = isotree::isolation.forest(X, ntrees=100L, sample_size=1024L,
									ndim=2, missing_action="fail",
									nthreads=1L)
}, times=7L)
```

```{r}
microbenchmark({
	iso = isotree::isolation.forest(X, ntrees=100L, sample_size=6435L,
									ndim=2, missing_action="fail",
									nthreads=1L)
}, times=7L)
```

```{r}
microbenchmark({
	iso = isotree::isolation.forest(X, ntrees=100L, sample_size=256L,
									ndim=2, missing_action="fail",
									nthreads=parallel::detectCores())
}, times=7L)
```

```{r}
microbenchmark({
	iso = isotree::isolation.forest(X, ntrees=100L, sample_size=1024L,
									ndim=2, missing_action="fail",
									nthreads=parallel::detectCores())
}, times=1L)
```

```{r}
microbenchmark({
	iso = isotree::isolation.forest(X, ntrees=100L, sample_size=6435L,
									ndim=2, missing_action="fail",
									nthreads=parallel::detectCores())
}, times=1L)
```

Timing IsolatonForest:
```{r}
microbenchmark({
	iso = IsolationForest::IsolationTrees(X_df, ntree=100L, rowSamp=TRUE, nRowSamp=256L)
}, times=7L)
```

```{r}
microbenchmark({
	iso = IsolationForest::IsolationTrees(X_df, ntree=100L, rowSamp=TRUE, nRowSamp=1024L)
}, times=7L)
```

```{r}
microbenchmark({
	iso = IsolationForest::IsolationTrees(X_df, ntree=100L, rowSamp=FALSE, nRowSamp=6435L)
}, times=7L)
```

Timing isofor:

```{r}
microbenchmark({
	iso = isofor::iForest(X, nt=100L, phi=256L)
}, times=1L)
```

```{r}
microbenchmark({
	iso = isofor::iForest(X, nt=100L, phi=1024L)
}, times=1L)
```

```{r}
microbenchmark({
	iso = isofor::iForest(X, nt=100L, phi=6435L)
}, times=1L)
```

Timing solitude:

```{r}
microbenchmark({
	iso = solitude::isolationForest$new(num_trees=100L, sample_size=256L,
										replace=FALSE, nproc=1L)
	sink("/dev/null")
	suppressMessages(iso$fit(X_df))
	sink()
}, times=7L)
```

```{r}
microbenchmark({
	iso = solitude::isolationForest$new(num_trees=100L, sample_size=1024L,
										replace=FALSE, nproc=1L)
	sink("/dev/null")
	suppressMessages(iso$fit(X_df))
	sink()
}, times=7L)
```

```{r}
microbenchmark({
	iso = solitude::isolationForest$new(num_trees=100L, sample_size=6435L,
										replace=FALSE, nproc=1L)
	sink("/dev/null")
	suppressMessages(iso$fit(X_df))
	sink()
}, times=7L)
```

```{r}
microbenchmark({
	iso = solitude::isolationForest$new(num_trees=100L, sample_size=256L,
										replace=FALSE, nproc=parallel::detectCores())
	sink("/dev/null")
	suppressMessages(iso$fit(X_df))
	sink()
}, times=7L)
```

```{r}
microbenchmark({
	iso = solitude::isolationForest$new(num_trees=100L, sample_size=1024L,
										replace=FALSE, nproc=parallel::detectCores())
	sink("/dev/null")
	suppressMessages(iso$fit(X_df))
	sink()
}, times=7L)
```

```{r}
microbenchmark({
	iso = solitude::isolationForest$new(num_trees=100L, sample_size=6435L,
										replace=FALSE, nproc=parallel::detectCores())
	sink("/dev/null")
	suppressMessages(iso$fit(X_df))
	sink()
}, times=7L)
```

### Mid-sized dataset: CovType (581,012 rows, 54 columns)

Dataset was taken from [LibSVM's webpage](https://www.csie.ntu.edu.tw/~cjlin/libsvmtools/datasets/).

```{r, message=FALSE, results=FALSE, warning=FALSE, comment=FALSE}
gc()
```
```{r}
X = readsparse::read.sparse("covtype.libsvm.binary")
X = as.matrix(X$X)
X_df = as.data.frame(X)
cat(dim(X))
```

******

Timing isotree, single-variable model:
```{r}
microbenchmark({
	iso = isotree::isolation.forest(X, ntrees=100L, sample_size=256L,
									ndim=1, missing_action="fail",
									nthreads=1L)
}, times=3L)
```

```{r}
microbenchmark({
	iso = isotree::isolation.forest(X, ntrees=100L, sample_size=1024L,
									ndim=1, missing_action="fail",
									nthreads=1L)
}, times=3L)
```

```{r}
microbenchmark({
	iso = isotree::isolation.forest(X, ntrees=100L, sample_size=10000L,
									ndim=1, missing_action="fail",
									nthreads=1L)
}, times=3L)
```

```{r}
microbenchmark({
	iso = isotree::isolation.forest(X, ntrees=100L, sample_size=256L,
									ndim=1, missing_action="fail",
									nthreads=parallel::detectCores())
}, times=3L)
```

```{r}
microbenchmark({
	iso = isotree::isolation.forest(X, ntrees=100L, sample_size=1024L,
									ndim=1, missing_action="fail",
									nthreads=parallel::detectCores())
}, times=3L)
```

```{r}
microbenchmark({
	iso = isotree::isolation.forest(X, ntrees=100L, sample_size=10000L,
									ndim=1, missing_action="fail",
									nthreads=parallel::detectCores())
}, times=3L)
```

Timing isotree, extended model:
```{r}
microbenchmark({
	iso = isotree::isolation.forest(X, ntrees=100L, sample_size=256L,
									ndim=2, missing_action="fail",
									nthreads=1L)
}, times=3L)
```

```{r}
microbenchmark({
	iso = isotree::isolation.forest(X, ntrees=100L, sample_size=1024L,
									ndim=2, missing_action="fail",
									nthreads=1L)
}, times=3L)
```

```{r}
microbenchmark({
	iso = isotree::isolation.forest(X, ntrees=100L, sample_size=10000L,
									ndim=2, missing_action="fail",
									nthreads=1L)
}, times=3L)
```

```{r}
microbenchmark({
	iso = isotree::isolation.forest(X, ntrees=100L, sample_size=256L,
									ndim=2, missing_action="fail",
									nthreads=parallel::detectCores())
}, times=3L)
```

```{r}
microbenchmark({
	iso = isotree::isolation.forest(X, ntrees=100L, sample_size=1024L,
									ndim=2, missing_action="fail",
									nthreads=parallel::detectCores())
}, times=3L)
```

```{r}
microbenchmark({
	iso = isotree::isolation.forest(X, ntrees=100L, sample_size=10000L,
									ndim=2, missing_action="fail",
									nthreads=parallel::detectCores())
}, times=3L)
```

IsolationForest runs out of memory and would take too long, not benchmarked.

isofor would take too long, not benchmarked.

Timing solitude:

```{r}
microbenchmark({
	iso = solitude::isolationForest$new(num_trees=100L, sample_size=256L,
										replace=FALSE, nproc=1L)
	sink("/dev/null")
	suppressMessages(iso$fit(X_df))
	sink()
}, times=1L)
```

```{r}
microbenchmark({
	iso = solitude::isolationForest$new(num_trees=100L, sample_size=1024L,
										replace=FALSE, nproc=1L)
	sink("/dev/null")
	suppressMessages(iso$fit(X_df))
	sink()
}, times=1L)
```

```{r}
microbenchmark({
	iso = solitude::isolationForest$new(num_trees=100L, sample_size=10000L,
										replace=FALSE, nproc=1L)
	sink("/dev/null")
	suppressMessages(iso$fit(X_df))
	sink()
}, times=1L)
```

```{r}
microbenchmark({
	iso = solitude::isolationForest$new(num_trees=100L, sample_size=256L,
										replace=FALSE, nproc=parallel::detectCores())
	sink("/dev/null")
	suppressMessages(iso$fit(X_df))
	sink()
}, times=1L)
```

```{r}
microbenchmark({
	iso = solitude::isolationForest$new(num_trees=100L, sample_size=1024L,
										replace=FALSE, nproc=parallel::detectCores())
	sink("/dev/null")
	suppressMessages(iso$fit(X_df))
	sink()
}, times=1L)
```

```{r}
microbenchmark({
	iso = solitude::isolationForest$new(num_trees=100L, sample_size=10000L,
										replace=FALSE, nproc=parallel::detectCores())
	sink("/dev/null")
	suppressMessages(iso$fit(X_df))
	sink()
}, times=1L)
```
